Leadership across cultures and contexts: an exploratory analysis

### 2020-05-13 Response to reviewers.Rmd

**There is little mention of gender in the Qualities of leadership section, even though this might be first attribute an alien might recognize as covarying with community-level leadership in human societies. Even in the more egalitarian societies, men often wield more overt political influence during camp or band-wide political discussions. Without getting into arguments about patriarchy and its origins, it might at least be useful to cite your own work on Chabu and work on the Tsimane (von Rueden et al. 2018) that finds similar traits predict leadership for men and women, implying its not gender per se that gives men a leadership advantage in politics.**

We now discuss sex-differences in community and political leadership, and refernce in greater deatils (albeit breif) the universal male bias in community leadership in the introuction and cite these works.

**How often are shamans also the most influential political figure (given your suggestion in line 317)? My sense from various ethnographies is that this was not the majority of cases (though often the case), whether in more egalitarian societies or chiefdoms. I could be wrong.**

We did not mean to imply shamans are often the most influential political leaders in this passage, only that managerial eliets (e.g., Stanish) are often associated with a special social status rooted in ideology. However, to this point we have now include a dedicated section in the results section to shaman leaders and directly address the reviewer's inquries. As we discuss, we see shamanism as a potentially important and overlooked leadership strategy.

### 2020-05-13 leadership_across_cultures_contexts.Rmd
Edits in conclusion.

We then investigated if the proportion of text records with evidence for *High status* differed among the different levels of group context, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastoralists were more likely to provide evidence for high status leaders, as were South America and East Eurasia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status leaders was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

@@

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominance and prestige styles of leadership. Our competencies cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* abilities dimension. @winkelman1987cross found that in 47% of societies sampled from the HRAF ($n = 43$) shamans were associated with providing decision-making services and associated with military power. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)).

<!-- We also found a link between prestige-based leadership, which is associated with deference towards accomplished individuals with expertise, special skills, and high intelligence [@cheng_dominance_2019; @henrich_evolution_2001], and charismatic leadership, which is associated with oratory skills, charismatic personality, confidence, and interpersonal skills [@grabo_charismatic_2016; @howell_role_2005]. See, Figure \@ref(fig:cluster)A, *Ability to provide social benefits* cluster. This suggests that these leadership strategies might be closely related across a broad range of societies and social contexts, or part of a broader strategy of social influence based on the provisioning of various benefits. -->

@@

These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and across all levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018].

<!-- IS THIS NEEDED? HERE ??? -->
These results provide a framework for further work on evolutionarily relevant costs and benefits of leadership, and suggests leaders-follower dynamics should continue to be prioritized in evolutionary models of demography, group dynamics, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016]

<!-- MAYBE IN CONCLUSION -->
Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural norms. Dichotomizing this complexity into two distinct social strategies, although theoretically and empirically convenient, likely fails to capture much of the ethnographic reality of human leadership dynamics.

@@

## Disentangling the relationship between leadership and high status, continued

These data present an opportunity to explore the relationship between high status and leadership, as represented in ethnographic texts. To do so, we draw on the term-document matrix of our ethnographic sample which was developed by computing the frequencies of all "informative" words in each text record after removing all punctuation and English stop words (uninformative words such as the, is, at), stemming the remaining words (reducing inflections such as plurals and past tenses to the root, or stem, words). In this data matrix the columns are all the unique word stems (terms), rows were the text records (documents), and the values in each cell were the frequencies of each word stem in each text record. The coding of *high status* was joined with this term-document matrix. This process produced a matrix with `r nrow(leader_dtm)` rows with one for each text-record and `r ncol(leader_dtm)` columns with a column for each unique term and a column for the researcher-coded *high status* value ($0$ or $1$).

We used lasso regression as a variable selection technique, which is especially useful when the number of predictors is large relative to the number of cases. Lasso regression has a penalty or shrinkage term, $\lambda$: larger values of $\lambda$ retain more variables in the model, at the risk of over-fitting, whereas smaller values of $\lambda$ decrease the number of predictors retained in the model, at the risk of under-fitting. The optimal value of $\lambda$ can be estimated with cross-validation [@tibshirani_regression_1996; @zou_regularization_2005]. We used the glmnet package [@friedman_regularization_2010] to fit a lasso regression model to determine which words best predicted evidence for high status as a leader quality. We used tenfold cross-validation to choose a $\lambda$ parameter that was within in standard error of the value that minimized the deviance. We then plotted all the non-zero coefficients (Figure \@ref(fig:highstatusplot)).

@@

Figure \@ref(fig:highstatusplot) depicts the `r length(highstatus_plot$data$Word)` words with non-zero (mostly positive) coefficients. The words most predictive of evidence for high status leaders are *`r rev(highstatus_plot$data$Word)[1:4]`*, and *`r rev(highstatus_plot$data$Word)[5]`*.

@@

@@

We constructed `r length(features)` "feature" variables from the higher-level clusters (Figure \@ref(fig:cluster)): `r paste(features, collapse = ', ')`. The Mediate, Prosociality, and Organize features were modestly positively correlated with each other, but not with Prestige or Social/reproductive/material success. Prosocial competencies was correlated with Cultural conformity. See Figure \@ref(fig:featurecor).

>>>>

We also investigated if the proportion of text records with evidence for *High status* differed among the different levels of group context, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were no clear differences. See Figure S\@ref(fig:groupstatus).

Hunter-gatherers and pastoralists were more likely to provide evidence for high status leaders, as were South America and East Eurasia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status leaders was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus).

@@

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominance and prestige styles of leadership. Our competencies cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* abilities dimension. @winkelman1987cross found that in 47% of societies sampled from the HRAF ($n = 43$) shamans were associated with providing decision-making services, a proscial and prestige-based computational service, and associated with military power, which is often associated with dominance. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)).

@@

These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018]. Given the broad range of leadership dimensions represented in the ethnographgric record, we suggest leaders-follower dynamics should be increasingly incorporated in evolutionary models beahvior and group dynamics, including demography, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016].

Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural norms. Dichotomizing this complexity into two distinct social strategies, although theoretically and empirically convenient, likely fails to capture much of the ethnographic reality of human leadership dynamics.

@@

## Predicting high status leaders with our four factors

<!-- These data present an opportunity to explore the relationship between high status and leadership, as represented in ethnographic texts. To do so, we draw on the term-document matrix of our ethnographic sample which was developed by computing the frequencies of all "informative" words in each text record after removing all punctuation and English stop words (uninformative words such as the, is, at), stemming the remaining words (reducing inflections such as plurals and past tenses to the root, or stem, words). In this data matrix the columns are all the unique word stems (terms), rows were the text records (documents), and the values in each cell were the frequencies of each word stem in each text record. The coding of *high status* was joined with this term-document matrix. This process produced a matrix with `r nrow(leader_dtm)` rows with one for each text-record and `r ncol(leader_dtm)` columns with a column for each unique term and a column for the researcher-coded *high status* value ($0$ or $1$).  -->

<!-- We used lasso regression as a variable selection technique, which is especially useful when the number of predictors is large relative to the number of cases. Lasso regression has a penalty or shrinkage term, $\lambda$: larger values of $\lambda$ retain more variables in the model, at the risk of over-fitting, whereas smaller values of $\lambda$ decrease the number of predictors retained in the model, at the risk of under-fitting. The optimal value of $\lambda$ can be estimated with cross-validation [@tibshirani_regression_1996; @zou_regularization_2005]. We used the glmnet package [@friedman_regularization_2010] to fit a lasso regression model to determine which words best predicted evidence for high status as a leader quality. We used tenfold cross-validation to choose a $\lambda$ parameter that was within in standard error of the value that minimized the deviance. We then plotted all the non-zero coefficients (Figure \@ref(fig:highstatusplot)).  -->

@@

<!-- Figure \@ref(fig:highstatusplot) depicts the `r length(highstatus_plot$data$Word)` words with non-zero (mostly positive) coefficients. The words most predictive of evidence for high status leaders are *`r rev(highstatus_plot$data$Word)[1:4]`*, and *`r rev(highstatus_plot$data$Word)[5]`*. -->

@@

As reported in the main text, we also investigated if the proportion of text records with evidence for *High status* differed among the different levels of group context, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were no clear differences and we interpret this as a null result. See Figure S\@ref(fig:groupstatus).

<!-- Hunter-gatherers and pastoralists were more likely to provide evidence for high status leaders, as were South America and East Eurasia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status leaders was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). -->

@@

We constructed `r length(features)` "feature" variables from the higher-level clusters (Figure \@ref(fig:cluster)): `r paste(features, collapse = ', ')`. The Strategize, Prosociality, and Organize features were modestly positively correlated with each other, but not with Prestige or Social/reproductive/material success. Prosocial competencies was correlated with Cultural conformity. See Figure \@ref(fig:featurecor).
